---
title: "FigS Bio-Oracle Temperature Data"
author: "Mary Fisher"
date: "2/1/2021"
output: html_document
---


# Description

Pull environmental data from [Bio-Oracle database](http://www.bio-oracle.org/) to explore gene x environment associations.

Originally written to extract environmental data for the US West Coast, edited in June 2020 to extract environmental data for the Korean peninsula (Yellow Sea, East Sea, Korea Strait).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This script requires the following packages
```{r}
library(here)
library(sdmpredictors)
library(leaflet)
library(tidyverse)
library(raster)
library(grid)
library(gridExtra)
```
<br>

# Download environmental layers 

You can explore the environmental data "layers" using the following code. It should return over 300 different environmental measurements.
```{r eval=FALSE, echo=TRUE}
layers.bio2 <- list_layers( datasets="Bio-ORACLE" )

## for temperature only
layers.bio2 %>% 
  filter(marine==TRUE & grepl('temperature',name)==TRUE) %>% 
  dplyr::select(name, layer_code)
```
<br>

Download the layer data for the metrics of interest: mean/max/range sea surface temperature, mean/min/max sea temperature at mean depth, mean/min/max sea temperature at min depth
```{r eval=FALSE, echo=TRUE}
environmental_vars <- load_layers( layercodes = c("BO_sstmean"), # salinity
                                   equalarea=FALSE, 
                                   rasterstack=TRUE)
```
<br>

Save a nice image of the mean annual SST around the Korean peninsula
```{r}
e <- extent(115,155,25,50)
plot(environmental_vars[[1]], col=hcl.colors(n=100, palette="viridis"),
     xlab="", ylab="", ext=e)

png('bio-oracle_sst_mean_zoom.png', height=400, width=400)
plot(environmental_vars[[2]], col=hcl.colors(n=100, palette="viridis"),
     xlab="", ylab="", ext=e)
dev.off()
```

















