---
title: "Bio Oracle Environmental Data"
author: "Natalie Lowell, Mary Fisher"
date: "Written 12/18/2019. Last Run `r Sys.time()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

# Description

Pull environmental data from [Bio-Oracle database](http://www.bio-oracle.org/) to explore gene x environment associations.

Originally written to extract environmental data for the US West Coast, edited in June 2020 to extract environmental data for the Korean peninsula (Yellow Sea, East Sea, Korea Strait).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This script requires the following packages
```{r}
library(here)
library(sdmpredictors)
library(leaflet)
library(tidyverse)
library(raster)
library(grid)
library(gridExtra)
```
<br>

# Download environmental layers 

You can explore the environmental data "layers" using the following code. It should return over 300 different environmental measurements.
```{r eval=FALSE, echo=TRUE}
layers.bio2 <- list_layers( datasets="Bio-ORACLE" )

## for all layers
layers.bio2 %>% 
  filter(marine==TRUE) %>% 
  dplyr::select(name, layer_code)

## for temperature only
layers.bio2 %>% 
  filter(marine==TRUE & grepl('temperature',name)==TRUE) %>% 
  dplyr::select(name, layer_code)
```
<br>

Download the layer data for the metrics of interest: mean/max/range sea surface temperature, mean/min/max sea temperature at mean depth, mean/min/max sea temperature at min depth
```{r eval=FALSE, echo=TRUE}
environmental_vars <- load_layers( layercodes = c("BO_sstmax", # max SST
                                                  "BO_sstmean", # mean SST
                                                  "BO_sstrange", # range SST
                                                  "BO2_tempmax_bdmean", # mean depth
                                                  "BO2_tempmean_bdmean",
                                                  "BO2_tempmin_bdmean", 
                                                  "BO2_templtmax_bdmean", #long-term
                                                  "BO2_tempmax_bdmax", # max depth
                                                  "BO2_tempmean_bdmax",
                                                  "BO2_tempmin_bdmax",
                                                  "BO2_templtmax_bdmax", #long-term
                                                  "BO_salinity"), # salinity
                                   equalarea=FALSE, 
                                   rasterstack=TRUE)
```
<br>

Save this as an R object to avoid downloading again!
```{r eval=FALSE, echo=TRUE}
# writeRaster(environmental_vars, filename=here::here('bio_oracle_tempvars.tif'), options="INTERLEAVE=BAND", overwrite=TRUE)
saveRDS(environmental_vars, here::here('input','bio_oracle_tempvars.rds'))
```
<br>

```{r include=FALSE}
## read back in environmental layers if this isn't the first time running the script
environmental_vars <- readRDS(here::here('input','bio_oracle_tempvars.rds'))
```
<br>

Check out the spatial extent of the bio-oracle data.
```{r}
par(mfrow=c(2,2))
for(i in seq(1,4)){image(environmental_vars, y=i)}
```
<br>

Let's zoom into the Korean peninsula.
```{r}
par(mfrow=c(2,2))
for(i in seq(1,4)){image(environmental_vars, y=i,xlim=c(100,150),ylim=c(25,55))}
```
<br>

Save a nice image of the mean annual SST around the Korean peninsula
```{r}
e <- extent(115,155,25,50)
plot(environmental_vars[[2]], col=hcl.colors(n=100, palette="viridis"),
     xlab="", ylab="", ext=e, zlim=c(0,27.5))
abline(v=seq(115,145,by=5), col="black", lty=3, lwd=1)
abline(h=seq(30,45,by=5), col="black", lty=3, lwd=1)
abline(h=c(25,50),col="black",lwd=2)
abline(v=c(110, 150), col="grey48", lty=3, lwd=1, cex=3)

png('bio-oracle_sst_mean_zoom.png', height=400, width=400)
plot(environmental_vars[[2]], col=hcl.colors(n=100, palette="viridis"),
     xlab="", ylab="", ext=e, zlim=c(0,27.5))
abline(v=seq(115,145,by=5), col="black", lty=3, lwd=1)
abline(h=seq(25,45,by=5), col="black", lty=3, lwd=1)
abline(v=c(110, 150), col="grey48", lty=3, lwd=1)
abline(h=c(25,50),col="black",lwd=1)
dev.off()
```
<br>


# Extract site data

These are the site locations for Pacific cod sampling around the Korean peninsula.
```{r echo=FALSE}
site_data <- data.frame(Sampling.Site = c("Block 161", "Boryeong", 
                                          "Namhae", "Geoje", "Jinhae Bay", "Pohang", 
                                          "Jukbyeon"),
           latitude= c(37,36.3,34.725,
                       34.8,35.05,36.1,
                       37.05),
           longitude = c(124.6,126.5,
                         127.836,128.7,128.7,129.5,
                         129.4))
site_data
```
<br>

Just grab the lat / lon coordinates
```{r}
my.sites.locs <- dplyr::select(site_data, latitude, longitude) %>%
  rename(Lon=longitude, Lat=latitude)

# Visualise sites of interest in google maps, make sure placing where we think we are
check_map <- leaflet()
check_map <- addTiles(check_map)
check_map <- addMarkers(check_map, lng=my.sites.locs$Lon, lat=my.sites.locs$Lat)
check_map
```
<br>


Extract environmental values from layers, stick into a dataframe
```{r}
my.sites.env.data <- data.frame(Name=site_data$Sampling.Site,
                                Var=raster::extract(environmental_vars, my.sites.locs))

str(my.sites.env.data)
```


# Average Site Data

```{r include=FALSE}
my.sites.env.data <- read.delim(here::here('mary_sites_env_data.tsv'), sep=" ")
```
<br>


`Bayenv` takes a single environmental measurement for each population. The environmental data layers were collected using sampling sites, with multiple sites making up a population. 
```{r echo=FALSE}
site_data <- site_data %>% mutate(population = ifelse(Sampling.Site %in% c("Block 161","Boryeong"), "West",
                                        ifelse(Sampling.Site == "Jukbyeon", "East", "South")),
                                  Sampling.Site = as.character(Sampling.Site))
site_data$Sampling.Site[which(site_data$Sampling.Site=="Block 161")] <- "Block161"
site_data$Sampling.Site[which(site_data$Sampling.Site=="Jinhae Bay")] <- "JinhaeBay"
site_data
```
<br>

Each row of my environmental data layers needs to be a specific environmental variable, and each column contains its value for a given population. 
```{r}
env.dat <- my.sites.env.data %>%
  left_join(dplyr::select(site_data,Sampling.Site,population),by=c('Name'='Sampling.Site')) %>%
  dplyr::select(-Name) %>%
  rename(Name=population) %>%
  dplyr::select(all_of(colnames(my.sites.env.data))) %>%
  group_by(Name) %>%
  summarise_all(mean,.groups='drop')
env.dat[,1:5]
```
<br>

Clean up column names, and re-order the populations so that they are in the same order as the genepop file.
```{r}
env.dat <- env.dat %>%
  clean_names() %>%
  rename(Name=name) %>%
  arrange(match(Name, c("South", "West", "East")))
colnames(env.dat)
```
<br>

Write out as a tab-delimited text file.
```{r}
write_delim(env.dat, 'mary_pops_environmental.txt', delim='\t')
```
<br>